\documentclass[12pt]{article}
\PassOptionsToPackage{hidelinks}{hyperref} 
\usepackage{amsmath}
\usepackage{tikz}
\usepackage{geometry}
\usepackage{bookmark}
\usepackage{graphicx}

\geometry{margin=2.5cm}

\title{N-Body simulation in OpenCl with a AMD iGPU}
\author{}
\date{}

\begin{document}

\begin{titlepage}
    \centering

    \vspace*{2cm}

    {\Huge \textbf{High-Performance N-Body Simulation}}\\[0.4cm]
    {\Large Using OpenCL on an AMD Integrated GPU}

    \vspace{2cm}

    {\Large \textbf{Nagy Levente}}

    \vspace{1cm}

    {\large \today}

    \vfill

    \rule{\textwidth}{0.4pt}

    \vspace{0.3cm}

    {\large
    Eötvös Lóránd Univeristy
    }

\end{titlepage}

\pagebreak

\tableofcontents


\pagebreak

\section{Introduction}

\subsection{N-Body simulation}

An N-body simulation models the motion of \textbf{N} particles in space,
where each particle is influenced by physical forces such as gravity,
defined by the gravitational constant \textbf{G}.

\section{Problem}
\subsection{General N-Body problem}

According to the nature of the space, no more than 2 bodies can orbit each other without colliding.
(Famously: 3-body problem)

\subsection{N-Body computing}
Computing the forces from each body acting on every other body is costly when \textbf{N} is a high number.
A simple brute force algorithm time complexity is \[O(n^2)\]
For a better algorithm exists the Barnes-Hut algorithm which time complexity is \[O(n \log(n))\]
So its a much better solution for computing the forces. We can plot them to see the difference

\subsection{Barnes-Hut alogithm}
\subsubsection{Idea}
The Barnes-Hut algorithm divides the space where the bodies located into 8 smaller chunks
if there is more than 1 body is in a chunk. After recursively doing this for every body
we are left with an Octree (each parent can have up to 8 children).
In the given tree each node represtens a 3Dimensional space with a body.
Than we compute the \textbf{COM} (Center Of Mass) of each node and use this formula
\[\theta > \frac{s}{d}\] 
where:
\begin{itemize}
    \item $s$ is the size of the region
    \item $d$ is the distance to the center of mass
    \item $\theta$ controls the trade-off between accuracy and performance
\end{itemize}
To determine if the nodes of the tree can be used as 1, reducing the computeitation steps we have to do reducing the time complexity
to just \[O(n*log(n))\]

\begin{center}
\begin{tikzpicture}
\node {Root}
child {node {Child 1}}
child {node {Child 2}}
child {node {Child 3}}
child {node {Child 4}}
child {node {Child 5}}
child {node {Child 6}}
child {node {Child 7}}
child {node {Child 8}};
\end{tikzpicture}
\end{center}

\subsubsection{Problem with the algorithm on GPU}
The Barnes-Hut algorithm relies on tree traversal, which is naturally recursive.
However, recursion is inefficient or unsupported in GPU programming models such as OpenCL.
Therefore, the algorithm must be reformulated using iterative techniques
and explicit memory structures.

\section{Solution}
This section is going to describe the solution and the implementation of the Barnes-Hut alogithm on the GPU
\subsection{Steps}
\begin{enumerate}
    \item Computing the bounding box containing all of the bodies
    \item Build the octree and instert the bodies into it
    \item Summerize the information of the bodies (Center Of Mass)
    \item Sorting the bodies based on spatial distance (not necessay but helps with the performance)
    \item Calculating the force
    \item Applying the force to each body, updating position and velocities
    \item Passing the buffer data to the Renderer 
\end{enumerate}

\subsection{Data Structure}

Efficient data representation is critical for achieving high performance on the GPU.
Instead of using a traditional pointer-based tree structure, the simulation uses a
\textbf{linearized octree} stored in contiguous memory buffers.

\paragraph{Overview}

All simulation data is stored in flat arrays in global memory.
Each index represents either a body or an internal node of the octree.

\begin{itemize}
    \item The first $N$ entries correspond to physical bodies (leaf nodes)
    \item The remaining entries represent internal octree nodes (parents and root)
\end{itemize}
The size of the array is chosen to be approximately $2N$, which provides a safe upper bound.
In an octree, each body corresponds to at most one leaf node, and the number of internal nodes
is strictly less than the number of bodies. Therefore, the total number of nodes
(bodies plus internal nodes) never exceeds $2N$.  

\subsection{Buffers}

\paragraph{Position and Mass Storage}

The positions and masses are stored in separate arrays:

\begin{itemize}
    \item \texttt{pos\_x, pos\_y, pos\_z} — coordinates of bodies and nodes
    \item \texttt{mass} — mass of each body or aggregated node
\end{itemize}

For internal nodes, the position represents the \textbf{center of mass (COM)}.

\paragraph{Tree Structure Representation}

The octree structure is encoded using index-based arrays:

\begin{itemize}
    \item \texttt{child} — stores indices of the 8 children for each node
    \item \texttt{start} — marks the beginning of a node's body range
\end{itemize}

Each node has exactly 8 child slots:

\[
\texttt{child[node * 8 + i]}, \quad i \in [0,7]
\]

Special values are used to indicate state:

\begin{itemize}
    \item \texttt{EMPTY (-1)} — no child present
    \item \texttt{LOCKED (-2)} — node is being modified by another thread
\end{itemize}

\paragraph{Node Properties}

Additional arrays store per-node metadata:

\begin{itemize}
    \item \texttt{nodeSize} — spatial size (half-width) of the node
    \item \texttt{nodeCount} — number of bodies contained in the node
\end{itemize}

The root node is stored at a fixed index (\texttt{NUMBER\_OF\_NODES}) and represents the bounding volume of the entire system.

\begin{itemize}
    \item \texttt{NUMBER\_OF\_NODES (numNodes)} — Size of te flat array so $2N$
\end{itemize}

\paragraph{Auxiliary Buffers}

Several additional buffers are used during computation:

\begin{itemize}
    \item \texttt{sorted} — stores bodies ordered by spatial locality
    \item \texttt{accX, accY, accZ} — accumulated forces
    \item \texttt{vx, vy, vz} — velocities of bodies
\end{itemize}

\paragraph{Advantages of the Approach}

This data-oriented design has several advantages:

\begin{itemize}
    \item Avoids dynamic memory allocation and recursion
    \item Enables parallel processing across thousands of threads
    \item Improves cache efficiency and memory coalescing
    \item Allows interoperability with OpenGL buffers for rendering
\end{itemize}

\paragraph{Summary}

By representing the octree as a set of flat arrays, the simulation achieves a structure
that is well-suited for GPU execution while still supporting efficient Barnes-Hut traversal.

\subsection{Kernels}

\subsubsection{Bounding Box Kernel}

The purpose of this kernel is to compute the global bounding box that contains all bodies in the simulation.
This bounding box is later used to initialize the root node of the octree structure required by the Barnes-Hut algorithm.

\paragraph{Overview}

The kernel performs a parallel reduction to determine the minimum and maximum coordinates along each axis ($x$, $y$, $z$).
Each work-group processes a subset of the bodies and computes a local bounding box, which is then combined into a global result.

\paragraph{Step 1: Local Reduction}

Each work-item iterates over the input data using a strided access pattern:

\[
i = \text{localId} + \text{groupId} \cdot \text{localSize}
\]

This ensures that all bodies are covered efficiently across all work-groups.

For each assigned body, the thread updates its local minimum and maximum values:
\begin{itemize}
    \item $x_{\min}, x_{\max}$
    \item $y_{\min}, y_{\max}$
    \item $z_{\min}, z_{\max}$
\end{itemize}

These values are stored in shared local memory to allow fast intra-group communication.

\paragraph{Step 2: Parallel Reduction in Local Memory}

A tree-based reduction is performed within each work-group:

\begin{itemize}
    \item At each step, half of the threads combine their values with a partner thread
    \item This continues until thread $0$ holds the result for the entire work-group
\end{itemize}

This reduces the per-group bounding box to a single set of values.

\paragraph{Step 3: Global Reduction}

The first thread of each work-group writes its result into global memory.
An atomic counter (\texttt{blockCount}) is used to detect when all work-groups have finished.
The last work-group to finish performs a final reduction over all partial results to compute the global bounding box.

\paragraph{Step 4: Root Node Initialization}

Once the global bounding box is known, the kernel initializes the root node of the octree:

\begin{itemize}
    \item The center of the root node is computed as:
    \[
    \mathbf{c} = \frac{1}{2} (\mathbf{min} + \mathbf{max})
    \]
    
    \item The size (radius) of the node is:
    \[
    r = \frac{1}{2} \max \left(
        x_{\max} - x_{\min},
        y_{\max} - y_{\min},
        z_{\max} - z_{\min}
    \right)
    \]
\end{itemize}

The root node is stored at index \texttt{NUMBER\_OF\_NODES} and initialized with:
\begin{itemize}
    \item Position = center of the bounding box
    \item Size = computed radius
    \item Mass = $-1$ (indicating an internal node)
    \item Children = \texttt{EMPTY}
\end{itemize}

\paragraph{Synchronization}

The kernel uses:
\begin{itemize}
    \item \texttt{barrier()} for synchronization within a work-group
    \item \texttt{atomic\_inc()} to coordinate between work-groups
\end{itemize}

This ensures correct ordering despite the lack of global synchronization in OpenCL.

\paragraph{Result}

At the end of execution:
\begin{itemize}
    \item The global bounding box is computed
    \item The root node of the octree is initialized
    \item The simulation is ready for tree construction
\end{itemize}

\end{document}